Tensor denoising and completion based on ordinal observations

### 2020-01-23 ICML/chanwoo/ICMLdraft.tex
Add files via upload

\section{Model}
\subsection{Low-rank cumulative model}
For the K-mode ordinal tensor $\mathcal{Y} = [\![ y_\omega]\!]\in[L]^{d_1\times\cdots\times d_K}$, we assume that its entries are realizations of independent multinomial random variables, such that
\begin{equation}
    \begin{aligned}
        \label{mainmodel}
        \mathbb{P}(y_\omega=l) = f_l(\theta_\omega), \quad \omega\in [d_1]\times\cdots\times[d_K].
    \end{aligned}
\end{equation}
In this model, a twice differentiable function $f_l : \mathbb{R}\rightarrow [0,1]$ with $l\in [L]$ is strictly increasing and satisfies $\sum_l f_l(\theta) = 1$ for a fixed $\theta$.
The tensor $\Theta = [\![\theta_\omega]\!]\in \mathbb{R}^{d_1\times\cdots\times d_K}$
is a hidden parameter which we are interested in. We assume the parameter tensor $\Theta$ is continuous value and admits a rank $\mr = (r_1,\cdots,r_K)$ Tucker decomposition,
\begin{equation}
    \begin{aligned}
        \label{tucker}
        \Theta = \mathcal{C}\times_1\mM_1\cdots\times_N\mM_N,
    \end{aligned}
\end{equation}
where $\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_K}$ is a core tensor and $\mM_k\in \mathbb{R}^{d_k\times r_k}$, for $n \in [L]$ is a factor matrix. We can get information about the influence of each mode by checking factor matrices.

@@

\begin{algorithm}[tb]
        \caption{Ordinal tensor decomposition }\label{alg}
        \begin{algorithmic}[]
            \STATE{\bfseries Input:}  \text{ Ordinal tensor }
            $\mathcal{Y}\in [L]^{d_1\times\cdots\times d_K}
            $, rank $\mr\in \mathbb{N_+}^{K-1}$, and entry-wise bound $\alpha\in \mathbb{R_+}$.
            \STATE{\bfseries Output:} $(\hat\Theta,\hat{\mb}) =  \argmax_{(\Theta,\mb)\in \mathcal{D}\times\mathcal{B}}  \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb).$
            \STATE Initialize $\mathcal{C}^{(0)}\in\mathbb{R}^{r_1\times\cdots\times r_k}, \{\mM_1^{(0)},\cdots,\mM_K^{(0)}\}$ and $\mb^{(0)} = ( b_1^{(0)},\cdots,b^{(0)}_{L-1})$ s.t. $b_1^{(0)}<\cdots<b_{L-1}^{(0)}$.
            \FOR{t = 1,2,$\cdots$,}
            \STATE{Update $\mM_k$}
            \FOR{k = 1,2,$\cdots$,K}
             \STATE $\mM_k^{(t+1)}\gets\argmax_{\mM_k\in\mathbb{R}^{d_k\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mM_k)$ given $ \mathcal{C}^{(t)},\{\mM_i^{(t)}: i\neq k\} \text{ and } \mb^{(t)}$ s.t.\\
             $\|\mathcal{C}^{(t)}\times_1\mM_1^{(t+1)}\cdots\times_k\mM_k^{(t+1)}\cdots\times_K\mM_K^{(t)}\|_\infty \leq \alpha$
            \ENDFOR
            \STATE {Update $\mathcal{C}$}
            \STATE $\mathcal{C}^{(t+1)} \gets \argmax_{\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C})$
            s.t.\newline $\|\mathcal{C}\times_1\mM_1^{(t+1)}\cdots\times_N\mM_K^{(t+1)}\|_\infty \leq \alpha$

            \STATE $\Theta^{(t+1)} \gets \mathcal{C}^{(t+1)}\times_1\mM_1^{(t+1)}\cdots\times_N\mM_K^{(t+1)}$
            \STATE \textbf{Update $\mb$}
            \STATE $\mb^{(t+1)} \gets \argmax_{\mb\in \mathbb{R}^{L-1}}\mathcal{L}_{\mathcal{Y},\Omega}\big(\Theta^{(t+1)},\mb\big)$  s.t. $b_1^{(0)}<\cdots<b_{L-1}^{(0)}$.
            \ENDFOR
            \STATE \textbf{return}
            $\Theta,\mb$
        \end{algorithmic}
    \end{algorithm}

@@

Also, we compare our ordinal tensor decomposition method with a logistic link function with the classical continuous Tucker decomposition method. We use 5 folded cross validation method for the comparison. Specifically, we randomly split the tensor entries into 5 similar sized pieces and alternatively use each piece of entries as a test data using other $80\%$ entries as a training data. The entries in the test data are encoded as missing and then predicted based on the tensor decomposition from the training data. We set the rank of the tensor as $\bm r = (23,23,8)$ which minimizes BIC value. Table \ref{CV} shows RMSE, MAE and error(false prediction rate), averaged over 5 test set results. We can check that our ordinal tensor decomposition model outperforms the classic continuous decomposition in all criteria.

\begin{table}[t]
\caption{Tensor completion for the HCP data. Two methods are compared: the proposed ordinal tensor decomposition and the classical continuous Tucker decomposition.}
\label{CV}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Method & RMSE & MAE & ERROR \\
\midrule
ORDINAL   & 0.1503711 & 0.1502662 & 0.1502151 \\
CONTI & 0.1603685& 0.1600219& 0.1598499\\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

>>>>

\section{Model}
\subsection{Low-rank cumulative model}
For the K-mode ordinal tensor $\mathcal{Y} = [\![ y_\omega]\!]\in[L]^{d_1\times\cdots\times d_K}$, we assume that its entries are realizations of independent multinomial random variables, such that
\begin{equation}
    \begin{aligned}
        \label{mainmodel}
        \mathbb{P}(y_\omega=l) = f_l(\theta_\omega), \quad \omega\in [d_1]\times\cdots\times[d_K].
    \end{aligned}
\end{equation}
In this model, a twice differentiable function $f_l : \mathbb{R}\rightarrow [0,1]$ with $l\in [L]$ is strictly increasing and satisfies $\sum_l f_l(\theta) = 1$ for a fixed $\theta$.
The tensor $\Theta = [\![\theta_\omega]\!]\in \mathbb{R}^{d_1\times\cdots\times d_K}$
is a hidden parameter which we are interested in. We assume the parameter tensor $\Theta$ is continuous value and admits a rank $\mr = (r_1,\cdots,r_K)$ Tucker decomposition,
\begin{equation}
    \begin{aligned}
        \label{tucker}
        \Theta = \mathcal{C}\times_1\mM_1\cdots\times_K\mM_N,
    \end{aligned}
\end{equation}
where $\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_K}$ is a core tensor and $\mM_k\in \mathbb{R}^{d_k\times r_k}$, for $n \in [L]$ is a factor matrix. We can get information about the influence of each mode by checking factor matrices.

@@

\begin{algorithm}[tb]
        \caption{Ordinal tensor decomposition }\label{alg}
        \begin{algorithmic}[]
            \STATE{\bfseries Input:}  \text{ Ordinal tensor }
            $\mathcal{Y}\in [L]^{d_1\times\cdots\times d_K}
            $,\\ \hspace{.43in} Rank $\mr\in \mathbb{N_+}^{K-1}$,\\
            \hspace{.43in} Entry-wise bound $\alpha\in \mathbb{R_+}$.
            \STATE{\bfseries Output:} $(\hat\Theta,\hat{\mb}) =  \argmax_{(\Theta,\mb)\in \mathcal{D}\times\mathcal{B}}  \mathcal{L}_{\mathcal{Y},\Omega}(\Theta,\mb).$
            \STATE Initialize  Core tensor $\mathcal{C}^{(0)}$,\\
            \hspace{.5in} Factor matrices $ \{\mM_1^{(0)},\cdots,\mM_K^{(0)}\}$,\\
            \hspace{.5in} Cut-off points $\mb^{(0)}$.
            \FOR{t = 1,2,$\cdots$,}
            \FOR{k = 1,2,$\cdots$,K}
            \STATE{Update $\mM_k$ fixing other blocks.}
            \STATE $\mM_k^{(t+1)}\gets\argmax_{\mM_k\in\mathbb{R}^{d_k\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mM_k)$ s.t.
             $\|\mathcal{C}^{(t)}\times_1\mM_1^{(t+1)}\cdots\times_k\mM_k\cdots\times_K\mM_K^{(t)}\|_\infty \leq \alpha$
            \ENDFOR
            \STATE {Update $\mathcal{C}$ fixing other blocks.}
            \STATE $\mathcal{C}^{(t+1)} \gets \argmax_{\mathcal{C}\in \mathbb{R}^{r_1\times\cdots\times r_k}}\mathcal{L}_{\mathcal{Y},\Omega}(\mathcal{C})$
            s.t.\newline $\|\mathcal{C}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}\|_\infty \leq \alpha$

            \STATE $\Theta^{(t+1)} \gets \mathcal{C}^{(t+1)}\times_1\mM_1^{(t+1)}\cdots\times_K\mM_K^{(t+1)}$
            \STATE {Update $\mb$ fixing $\Theta^{(t+1)}$.}
            \STATE $\mb^{(t+1)} \gets \argmax_{\mb\in \mathbb{R}^{L-1}}\mathcal{L}_{\mathcal{Y},\Omega}\big(\Theta^{(t+1)},\mb\big)$
            \ENDFOR
            \STATE \textbf{return}
            $\Theta,\mb$
        \end{algorithmic}
    \end{algorithm}

@@

Also, we compare our ordinal tensor decomposition method with a logistic link function with other methods: the continuous tensor decomposition method, the 1 bit-tensor completion method with probit link function and the 1 bit-tensor completion method with logistic link function. We use 5 folded cross validation method for the comparison. Specifically, we randomly split the tensor entries into 5 similar sized pieces and alternatively use each piece of entries as a test data using other $80\%$ entries as a training data. The entries in the test data are encoded as missing and then predicted based on each method from the training data. We set the rank of the tensor as $\bm r = (23,23,8)$ for the tensor decomposition methods which minimizes BIC value. Table \ref{CV} shows RMSE, MAE and error(false prediction rate), averaged over 5 test set results. We can check that our ordinal tensor decomposition model outperforms the other methods in all criteria.

\begin{table}[t]
\label{CV}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
 Method & RMSE & MAE & ERROR \\
\midrule
\begin{tabular}[c]{@{}c@{}}Oridnal tensor\\ Decomposition\end{tabular}  & 0.1503711 & 0.1502662 & 0.1502151 \\ \\
\begin{tabular}[c]{@{}c@{}}Conti tensor\\ Decomposition\end{tabular} & 0.1603685& 0.1600219& 0.1598499\\ \\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (probit)\end{tabular}  & 0.3658390 & 0.3632757 & 0.3619940\\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (logistic)\end{tabular}  &  0.4519253 & 0.4218513 & 0.4068143 
\\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\caption{Results of comparisons among 4 methods on the HCP data predicting the test data. Four methods are the ordinal tensor decomposition algorithm, the continuous tensor decomposition algorithm and the 1 bit tensor completion method with probit link function and logistic link function. Each method is evaluated by RMSE, MAE, error(false prediction rate).}
\end{table}
